---
title: "Canopy Segmentation Explorer"
author: "John Sulik"
format: html
server: shiny
---

Upload a drone RGB orthomosaic and a plot-boundary file, explore the
vegetation index and threshold live, then process at full resolution to get
a vegetation mask, a soil-masked RGB image, and percent cover per plot.

**Typical workflow:**

1. Upload your orthomosaic and plot boundaries (**1. Files**).
2. Sparse or early-emergence canopy can be hard to discern over a whole
   field so if that's an issue, turn on **Zoom preview to one plot**
   (**2. Preview area**).
3. If your orthomosaic includes a near-infrared band, turn that on and set
   which bands are NIR and red (**3. Near-infrared**) to make NDVI an option.
4. Pick an index (**4. Index**), click the (i) next to it if you're not
   sure which one to use or why its scale looks the way it does, then drag
   the threshold slider next to the histogram until the segmentation
   overlay matches what you'd call vegetation by eye.
5. Click **Process at full resolution** (**5. Full-resolution export**) to
   compute the real vegetation mask, soil-masked RGB image, and per-plot
   cover, and download them.

```{r}
#| context: setup

invisible(capture.output(suppressPackageStartupMessages({
  library(shiny)
  library(terra)
})))

terraOptions(progress = 0, memfrac = 0.3, todisk = TRUE)
options(shiny.maxRequestSize = 3000 * 1024^2)  # drone orthomosaics run large; default Shiny limit is 5MB

PREVIEW_MAX_DIM <- 500  # longest side, in px, of the downsampled preview everything below the file inputs reacts to

norm_fun <- function(r, g, b) {
  # Normalized ExGR, following Woebbecke et al. (1995): normalize each band to
  # its share of total brightness first (r+g+b=1), then combine as 3g-2.4r-b.
  # Guarded against dividing by zero for a pure-black pixel.
  denom <- r + g + b
  rn <- ifelse(denom > 0, r / denom, NA)
  gn <- ifelse(denom > 0, g / denom, NA)
  bn <- ifelse(denom > 0, b / denom, NA)
  3 * gn - 2.4 * rn - bn
}
raw_fun <- function(r, g, b) {
  # Raw ExGR = ExG - ExR = (2G - R - B) - (1.4R - G) = 3G - 2.4R - B, unnormalized.
  3 * g - 2.4 * r - b
}
grvi_fun <- function(r, g, b) {
  # Green-Red Vegetation Index: (G - R) / (G + R) -- same normalized-difference
  # form as NDVI, but built entirely from visible bands.
  denom <- g + r
  ifelse(denom != 0, (g - r) / denom, NA)
}
exg_fun <- function(r, g, b) {
  # Excess Green: 2G - R - B, unnormalized (same large scale as raw ExGR).
  2 * g - r - b
}
ndvi_fun <- function(nir, red) {
  denom <- nir + red
  ifelse(denom != 0, (nir - red) / denom, NA)
}

# One entry per index that only needs the orthomosaic's own R/G/B bands.
# NDVI is kept separate below since it needs a different pair of bands (NIR
# + red, picked by the user) and is only ever available once nir_band is
# set -- everything that varies per index (its formula, default threshold,
# and the explanation behind its info icon) lives here once, rather than
# repeated across the preview, threshold-default, and full-resolution-export
# logic further down.
rgb_indices <- list(
  "Normalized ExGR" = list(
    fun = norm_fun, default = -0.01,
    info = "r = R/(R+G+B), g = G/(R+G+B), b = B/(R+G+B); ExG - ExR = (2g - r - b) - (1.4r - g) = 3g - 2.4r - b"
  ),
  "Raw ExGR" = list(
    fun = raw_fun, default = 0,
    info = "ExG - ExR = (2G - R - B) - (1.4R - G) = 3G - 2.4R - B"
  ),
  "GRVI" = list(
    fun = grvi_fun, default = 0,
    info = "(G - R) / (G + R)"
  ),
  "ExG" = list(
    fun = exg_fun, default = 0,
    info = "2G - R - B"
  )
)
ndvi_info <- "(NIR - Red) / (NIR + Red)"

default_threshold_for <- function(index_method) {
  if (identical(index_method, "NDVI")) return(0.2)
  spec <- rgb_indices[[index_method]]
  if (!is.null(spec)) spec$default else 0
}

# input id for each index's (i) info button.
index_info_id <- c("Normalized ExGR" = "info_norm_exgr", "Raw ExGR" = "info_raw_exgr",
                    "GRVI" = "info_grvi", "ExG" = "info_exg", "NDVI" = "info_ndvi")

available_indices <- function(include_ndvi) c(names(rgb_indices), if (include_ndvi) "NDVI")

# Hand-built rather than radioButtons(choiceNames=...): reproduces the same
# markup radioButtons() itself generates (each option is a div.radio >
# label > input[type=radio] + span, all inside a div.shiny-input-radiogroup
# with id "index_method"), so Shiny's normal client-side binding still
# picks it up as a working input, but puts each option's (i) button as a
# SIBLING of the label instead of nested inside it. Nesting a button inside
# the label would double as clicking the radio itself (labels toggle their
# associated input on any click within them), and separately, an actionLink
# placed there did not register clicks reliably at all when embedded in
# radioButtons()'s own choiceNames and later swapped via
# updateRadioButtons() -- confirmed directly, both a synthetic and a real
# simulated click left its counter at 0.
index_radio_ui <- function(nm, selected) {
  tags$div(id = "index_method", class = "form-group shiny-input-radiogroup shiny-input-container",
           role = "radiogroup", `aria-labelledby` = "index_method-label",
    tags$label(class = "control-label", id = "index_method-label", `for` = "index_method", "Index"),
    tags$div(class = "shiny-options-group",
      lapply(nm, function(x) {
        tags$div(class = "radio",
          tags$label(
            tags$input(type = "radio", name = "index_method", value = x,
                       checked = if (identical(x, selected)) "checked" else NULL),
            tags$span(x)
          ),
          actionButton(index_info_id[[x]], label = icon("circle-info"), class = "btn-link btn-xs",
                       style = "padding: 0 0 0 6px; vertical-align: baseline; line-height: 1;")
        )
      })
    )
  )
}

# Blends vegetation-classified cells toward a highlight color directly in
# the R/G/B bands rather than layering a semi-transparent mask on top --
# terra's own plot(mask, col=c(NA,"green"), add=TRUE) does not composite
# transparently over plotRGB() (an NA-mapped cell paints opaque white,
# hiding the photo underneath), discovered by looking at a rendered test
# image, not by reading the code. Red rather than green specifically:
# green-on-green canopy makes it hard to spot misclassified pixels: red
# reads clearly against both green vegetation and tan/pink soil.
blend_overlay <- function(rgb_rast, idx_rast, threshold) {
  is_veg <- idx_rast >= threshold
  alpha <- 0.45
  highlight <- c(230, 25, 25)
  r2 <- ifel(is_veg, (1 - alpha) * rgb_rast[[1]] + alpha * highlight[1], rgb_rast[[1]])
  g2 <- ifel(is_veg, (1 - alpha) * rgb_rast[[2]] + alpha * highlight[2], rgb_rast[[2]])
  b2 <- ifel(is_veg, (1 - alpha) * rgb_rast[[3]] + alpha * highlight[3], rgb_rast[[3]])
  out <- c(r2, g2, b2)
  names(out) <- c("R", "G", "B")
  out
}

# Shiny renames every uploaded file to a meaningless temp name (0.dat,
# 1.dat, ...), which breaks multi-file formats like shapefiles whose parts
# (.shp/.dbf/.shx/.prj) must sit next to each other under matching names.
# Copies each upload into one temp folder under its ORIGINAL name instead,
# and returns the path to whichever file terra should actually open.
stage_upload <- function(file_input, pattern) {
  dir <- tempfile("upload_")
  dir.create(dir)
  file.copy(file_input$datapath, file.path(dir, file_input$name))
  main <- file_input$name[grepl(pattern, file_input$name, ignore.case = TRUE)]
  req(length(main) >= 1)
  file.path(dir, main[1])
}
```

```{r}
#| label: layout

tagList(

tags$style(HTML("
  .busy-spinner { display: none; align-items: center; gap: 8px; margin: 4px 0 10px 0; color: #666; }
  .shiny-busy .busy-spinner { display: flex; }
  .busy-spinner .spin {
    width: 16px; height: 16px; flex: none;
    border: 2px solid #ccc; border-top-color: #666; border-radius: 50%;
    animation: busy-spin 0.8s linear infinite;
  }
  @keyframes busy-spin { to { transform: rotate(360deg); } }
")),

sidebarLayout(

  sidebarPanel(
    width = 4,

    h4("1. Files"),
    fileInput("ortho_file", "Drone orthomosaic (GeoTIFF)", accept = c(".tif", ".tiff")),
    fileInput("plots_file", "Plot boundaries", multiple = TRUE,
              accept = c(".geojson", ".json", ".gpkg", ".shp", ".dbf", ".shx", ".prj")),
    helpText("For a shapefile, select all of .shp/.dbf/.shx/.prj together."),

    h4("2. Preview area"),
    checkboxInput("zoom_plot", "Zoom preview to one plot", value = FALSE),
    conditionalPanel(
      "input.zoom_plot",
      selectInput("zoom_plot_id", "Plot", choices = character(0))
    ),
    helpText("Averaging over the whole field can wash out sparse or early-emergence canopy. Zooming to about a third of one plot's area keeps real detail visible instead of blending it away. This only affects what you see below, never the full-resolution export."),

    h4("3. Near-infrared (optional)"),
    checkboxInput("use_ndvi", "My orthomosaic has a near-infrared band", value = FALSE),
    conditionalPanel(
      "input.use_ndvi",
      numericInput("nir_band", "NIR band number", value = 4, min = 1, step = 1),
      numericInput("red_band", "Red band number", value = 1, min = 1, step = 1)
    ),

    h4("4. Index"),
    uiOutput("index_method_ui"),
    helpText("Drag the threshold slider next to the histogram below."),
    checkboxInput("show_overlay", "Show segmentation overlay", value = TRUE),

    h4("5. Full-resolution export"),
    actionButton("process", "Process at full resolution", class = "btn-primary"),
    br(), br(),
    uiOutput("download_ui")
  ),

  mainPanel(
    width = 8,
    h5("Segmentation overlay"),
    div(class = "busy-spinner", div(class = "spin"), span("Computing, this can take a moment for a large orthomosaic...")),
    plotOutput("overlay_plot", height = "350px"),
    h5("RGB"),
    plotOutput("rgb_plot", height = "350px"),
    h5("Histogram and threshold"),
    plotOutput("hist_plot", height = "250px"),
    sliderInput("threshold", "Threshold", min = -1, max = 1, value = -0.01, step = 0.001, width = "100%"),
    helpText("If the overlay is missing plant pixels you want included, push the slider left (lower threshold). If it's picking up too much soil, push it right (raise threshold)."),
    strong(textOutput("veg_fraction_text")),
    h4("Vegetation cover by plot"),
    p("Only populated after Process at full resolution. This table (and the downloads above) use the exact, full-resolution mask, not the preview."),
    tableOutput("plot_cover_table")
  )
)

)
```

```{r}
#| context: server

# --- Load whatever was uploaded ------------------------------------------

ortho_full <- reactive({
  req(input$ortho_file)
  rast(input$ortho_file$datapath)
})

rgb <- reactive({
  s <- ortho_full()[[1:3]]
  names(s) <- c("R", "G", "B")
  s
})

plots_vec <- reactive({
  req(input$plots_file)
  path <- stage_upload(input$plots_file, "\\.(geojson|json|gpkg|shp)$")
  v <- vect(path)
  if (!same.crs(v, rgb())) v <- project(v, rgb())  # keep zonal stats meaningful if the plot file arrives in a different projection
  v
})

# "Plot 3 (D)" when the plot file carries a first attribute worth showing (e.g. Treatment), else just "Plot 3".
observeEvent(plots_vec(), {
  pv <- plots_vec()
  df <- as.data.frame(pv)
  n <- nrow(pv)
  labels <- if (ncol(df) >= 1) paste0("Plot ", seq_len(n), " (", df[[1]], ")") else paste0("Plot ", seq_len(n))
  updateSelectInput(session, "zoom_plot_id", choices = setNames(seq_len(n), labels))
})

# --- Downsampled preview: fast, rebuilt whenever the files, NIR/red band, or preview area change ---

# Either the full orthomosaic, or (when "Zoom preview to one plot" is on) a
# window sized to a third of that plot's own area, centered on it -- so a
# thin or sparse early-emergence canopy doesn't get averaged away over the
# whole field. Only feeds the preview below; the full-resolution export
# always uses the complete image regardless of this setting.
preview_source <- reactive({
  full <- ortho_full()
  if (!isTRUE(input$zoom_plot)) return(full)
  req(input$zoom_plot_id, plots_vec())
  idx <- as.integer(input$zoom_plot_id)
  pv <- plots_vec()
  req(idx >= 1, idx <= nrow(pv))
  one_plot <- pv[idx, ]
  side <- sqrt(expanse(one_plot) / 3)
  ctr <- crds(centroids(one_plot, inside = TRUE))[1, ]
  crop(full, ext(ctr[1] - side / 2, ctr[1] + side / 2, ctr[2] - side / 2, ctr[2] + side / 2))
})

preview <- reactive({
  full_p <- preview_source()
  s <- full_p[[1:3]]
  names(s) <- c("R", "G", "B")
  fact <- max(1L, ceiling(max(nrow(s), ncol(s)) / PREVIEW_MAX_DIM))
  preview_rgb <- aggregate(s, fact = fact, fun = "mean", na.rm = TRUE)

  ndvi_prev <- if (isTRUE(input$use_ndvi)) {
    req(input$nir_band, input$red_band)
    nir_p <- aggregate(full_p[[input$nir_band]], fact = fact, fun = "mean", na.rm = TRUE)
    red_p <- aggregate(full_p[[input$red_band]], fact = fact, fun = "mean", na.rm = TRUE)
    lapp(c(nir_p, red_p), ndvi_fun)
  } else {
    NULL
  }

  result <- lapply(rgb_indices, function(spec) lapp(preview_rgb, spec$fun))
  result$rgb <- preview_rgb
  result$NDVI <- ndvi_prev
  result
})

# Index choices grow an NDVI option only once it's actually available; the
# whole radio group is rebuilt (see index_radio_ui() in Setup) so each
# option keeps its own (i) button next to it.
output$index_method_ui <- renderUI({
  nm <- available_indices(isTRUE(input$use_ndvi))
  selected <- if (isTruthy(input$index_method) && input$index_method %in% nm) input$index_method else nm[1]
  index_radio_ui(nm, selected)
})

observeEvent(input$info_norm_exgr, showModal(modalDialog(title = "Normalized ExGR", rgb_indices[["Normalized ExGR"]]$info, easyClose = TRUE)))
observeEvent(input$info_raw_exgr, showModal(modalDialog(title = "Raw ExGR", rgb_indices[["Raw ExGR"]]$info, easyClose = TRUE)))
observeEvent(input$info_grvi, showModal(modalDialog(title = "GRVI", rgb_indices[["GRVI"]]$info, easyClose = TRUE)))
observeEvent(input$info_exg, showModal(modalDialog(title = "ExG", rgb_indices[["ExG"]]$info, easyClose = TRUE)))
observeEvent(input$info_ndvi, showModal(modalDialog(title = "NDVI", ndvi_info, easyClose = TRUE)))

active_index <- reactive({
  req(input$index_method)
  preview()[[input$index_method]]
})

# Threshold slider tracks whichever index is active: its own real data range, and a sensible default.
observeEvent(active_index(), {
  idx <- active_index()
  req(idx)
  r <- range(values(idx)[, 1], na.rm = TRUE)
  if (!all(is.finite(r))) return()
  updateSliderInput(session, "threshold",
                     min = r[1], max = r[2],
                     value = default_threshold_for(input$index_method),
                     step = (r[2] - r[1]) / 500)
})

# --- Preview plots and stats: instant, pure R-side reactivity, no re-render step ---

output$rgb_plot <- renderPlot({
  req(preview())
  plotRGB(preview()$rgb, r = 1, g = 2, b = 3, stretch = "lin", mar = c(0, 0, 0, 0))
})

output$overlay_plot <- renderPlot({
  idx <- active_index()
  req(idx)
  if (isTRUE(input$show_overlay)) {
    plotRGB(blend_overlay(preview()$rgb, idx, input$threshold), r = 1, g = 2, b = 3, stretch = "lin", mar = c(0, 0, 0, 0))
  } else {
    plotRGB(preview()$rgb, r = 1, g = 2, b = 3, stretch = "lin", mar = c(0, 0, 0, 0))
  }
})

output$hist_plot <- renderPlot({
  idx <- active_index()
  req(idx)
  v <- values(idx)[, 1]
  h <- hist(v, breaks = 60, plot = FALSE)
  # Bars at or above the threshold are what the overlay currently counts as vegetation.
  bar_col <- ifelse(h$mids >= input$threshold, "forestgreen", "grey80")
  plot(h, col = bar_col, border = "white", main = NULL, xlab = input$index_method, ylab = "Count")
  abline(v = input$threshold, col = "red", lwd = 2)
})

output$veg_fraction_text <- renderText({
  idx <- active_index()
  req(idx)
  v <- values(idx)[, 1]
  frac <- 100 * mean(v >= input$threshold, na.rm = TRUE)
  suffix <- if (isTRUE(input$show_overlay)) "" else "  (overlay hidden)"
  sprintf("Vegetation fraction (preview): %.2f%%%s", frac, suffix)
})

# --- Full-resolution export: explicit, slow, only runs when asked ---

full_res <- eventReactive(input$process, {
  req(input$plots_file)
  withProgress(message = "Processing at full resolution", value = 0, {

    incProgress(0.1, detail = "computing index")
    index_full <- if (input$index_method == "NDVI") {
      req(input$nir_band, input$red_band)
      lapp(c(ortho_full()[[input$nir_band]], ortho_full()[[input$red_band]]), ndvi_fun)
    } else {
      lapp(rgb(), rgb_indices[[input$index_method]]$fun)
    }

    incProgress(0.4, detail = "thresholding vegetation mask")
    mask_path <- tempfile(fileext = ".tif")
    vf_mask <- ifel(index_full >= input$threshold, 1, 0,
                     filename = mask_path, overwrite = TRUE, wopt = list(datatype = "INT1U"))
    names(vf_mask) <- "veg_fraction"

    incProgress(0.2, detail = "soil-masked RGB")
    soiless_path <- tempfile(fileext = ".tif")
    mask(rgb(), vf_mask, maskvalues = 0,
         filename = soiless_path, overwrite = TRUE, wopt = list(datatype = "INT1U"))

    incProgress(0.2, detail = "vegetation cover by plot")
    plots_v <- plots_vec()
    # rasterize()+zonal() rather than terra::extract(plots, ...): for long,
    # thin plots (strip trials running nearly the full length of a field),
    # extract()'s per-polygon clipping can exhaust memory on a large
    # raster -- confirmed directly on a real 6-strip field. Rasterizing the
    # plot IDs onto the mask's own grid once, then computing zonal means,
    # gets identical numbers without ever clipping a polygon out of the
    # raster.
    zone_id <- rasterize(plots_v, vf_mask, field = seq_len(nrow(plots_v)))
    names(zone_id) <- "zone"
    veg_zonal <- zonal(vf_mask, zone_id, fun = "mean", na.rm = TRUE)
    zone_match <- match(seq_len(nrow(plots_v)), veg_zonal$zone)
    plots_v$PctCover <- round(100 * veg_zonal$veg_fraction[zone_match], 1)

    incProgress(0.1)
    list(mask_path = mask_path, soiless_path = soiless_path, plots_df = as.data.frame(plots_v))
  })
})

output$plot_cover_table <- renderTable({
  full_res()$plots_df
})

output$download_ui <- renderUI({
  req(full_res())
  tagList(
    downloadButton("dl_mask", "Vegetation mask (.tif)"),
    downloadButton("dl_soiless", "Soil-masked RGB (.tif)"),
    downloadButton("dl_table", "Plot cover (.csv)")
  )
})

output$dl_mask <- downloadHandler(
  filename = function() "vegetation_mask.tif",
  content = function(file) file.copy(full_res()$mask_path, file)
)
output$dl_soiless <- downloadHandler(
  filename = function() "soil_masked_rgb.tif",
  content = function(file) file.copy(full_res()$soiless_path, file)
)
output$dl_table <- downloadHandler(
  filename = function() "vegetation_cover_by_plot.csv",
  content = function(file) write.csv(full_res()$plots_df, file, row.names = FALSE)
)
```
